---
title: "Adding data to the BEH Walk SmartMap"
date: "2025-11-25"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      eval = FALSE)
#install.packages("DT", type = "binary")
```
```{r load functions, eval=TRUE, echo=FALSE, message = FALSE}

invisible(
  capture.output(
    suppressMessages(
    suppressWarnings({

list.of.packages <- c('dplyr', 'glue', 'vctrs', 'DT', 'tidycensus', 'terra')
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages,function(x){library(x,character.only=TRUE)}) 
library(glue)
library('RPostgres')
library(tidyr)
## UPDATE Working directory
wd <- 'H:/projects/beh_nwi_smartmap'
setwd('H:/projects/beh_nwi_smartmap')
getwd()
source(glue('{wd}/tools/create_walkability_measures.R'))
source(glue('{wd}/tools/walkdb_setup.R'))

## clear tigris cache before downloading:
unlink(tigris::tigris_cache_dir('~/.tigris_cache'), recursive = TRUE)
tigris::tigris_cache_dir("~/.tigris_cache")
options(tigris_use_cache = FALSE)
readRenviron('~/.Renviron')

# connect to database & set dsn
## update database username & password
U <- us_census <- connectdb(dbname = 'nwi_smartmap', host = 'doyenne.csde.washington.edu', port = 5432, 
                   password = 'csde4322', user = 'aybloom')
dsn = "PG:host=doyenne.csde.washington.edu dbname=nwi_smartmap user=aybloom password=csde4322 port=5432" 

})
)))

```

# Check database schemas and data availability for years / states
*note that water data is not available for years prior to 2011*

  List current schemas and database tables set up in the nwi_smartmap database
  <!-- Should include: (1) tract land, (2) area water, (3) tract_nowater, (4) tract_variables, and (4) edges (intersections) -->

```{r check db, eval = TRUE}
# Ensure schemas exist
check_for_schemas()
# check for data by state / year
state_year_list <- get_db_list_states_years()

state_year_list %>% arrange(statefp, year)

```

## Check raster file availability for years / states

  List of the raster files that have been created so far for each year. 
  
  Raster files should include: (1) land area, (2) population density, (3) intersection density, and (4) business density.
  
```{r list raster files, eval = TRUE}
check_raster_files()
```

### Add data for set states and years to Smartwalk database
```{r get missing data, eval=FALSE}
# set years / states list 
year_list <- c(2007)
state_abbr_list <- c('pa', 'mn')

# set up database partitions for years / states to add data
f_setup_database(conn = U, year_list = year_list, state_abbr_list = state_abbr_list)

# download and add tract land, water, land with water erased, population variables and edges to database
f_add_data_smartwalk_db(conn = U, years=year_list, state_abbr_list = state_abbr_list)



```

## Create state-level raster files for land area, population, intersection density

```{r rasterize intersection, land, population}

# Create raster files for tract land area, population density, intersection density 
f_create_raster_files(conn = U, years = year_list, state_abbr_list = state_abbr_list)
```

## Create state-level raster files for business density

To get business data, you need to access the data enclave (see data enclave r code folder)
** First transfer the 2016 land area file for the state of interest to the data enclave folder (to ensure consistent extents of the raster file). 

** Next, use the code in the r_code_base folder while in the data enclave to create a raster file of businesses for a given year and state. The get_raster_stacks.r file creates and runs the function to create a raster. 

** To export the raw raster file, use the Data Enclave process to export from the data enclave and save the state_raster/business/{yr}/sum folder to run the function below. 

```{r rasterize business data}
 state_fips <- fips_codes %>% filter(state %in% toupper(state_abbr_list))
 for(state_fips_code in unique(state_fips$state_code)){
    for(yr in years){
      business_sum_filepath <- glue('H:/projects/beh_nwi_smartmap/data/state_raster/business/{yr}/sum/state_raster_business_density_{yr}_{state_fips_code}.tif')
      if(file.exists(business_sum_filepath)){
      f_get_business_density(state_fips = state_fips_code,
                             year = yr,
                             verbose = TRUE)
      } else if(!file.exists(business_sum_filepath)){
       message(glue('.... need to extract business data for {year}, for {state} from the Data Enclave'))
      }
    }}
```

## Create a composite walk measures
The Composite walk measure is created by finding the year- and state- specific z-score of (1) businesses density, (2) population density, and (3) street network density and summing to create a composite measure. The function below calcualtes the z-score for each state/year and creates a raster files with the composite scores for a given year/state. 

```{r create composite scores}
 state_fips <- fips_codes %>% filter(state %in% toupper(state_abbr_list))
for(yr in year_list){
  for(f in state_fips){

# create composite walk scores
f_create_composite_measure_raster(yr = yr, 
                                              state_fips = f)
}}
```

