---
title: "Walkability data pulls for longitudinal data"
author: "Amy Youngbloom"
date: "2025-09-24"
output: html_document
---

```{r setup, include=FALSE}
library(RPostgres)
library(DBI)
library(leaflet)
library('glue')
library(sf)
```
## connect the database to R environment
```{r connect db}
# PostGIS db (docker)
source("H:/projects/beh_nwi_smartmap/dbconnect.R")
U <- us_census <- connectdb()
```

## get tract functions download post gis partition functions & set up the tables and partitions
```{r tract functions}
source('tools/tracts_download_postgis_partition.R')
# update states and fips: 
# states and counties FIPS codes
states_fips <- '53'
counties_fips <- fips_codes %>% dplyr::filter(state_code %in% states_fips)
dsn = "PG:host=doyenne.csde.washington.edu dbname=nwi_smartmap user=aybloom password=csde4322 port=5432"

#update years: 
years<-c(2016)

# UPDATE - add create tmp schema to prevent error thrown
dbExecute(U, "CREATE SCHEMA IF NOT EXISTS tmp;")
# create schemas
f_tract_create_schema()
 ## throws error: tmp.tract does not exist at character 46
#create table for single tract
T()

# create schemas
f_tract_create_partitions()

```

```{r view all the schemas and tables and partitions}
parents <- dbGetQuery(U, "
  SELECT schemaname
  FROM pg_catalog.pg_tables t
  WHERE t.schemaname NOT IN ('pg_catalog', 'information_schema')
    AND t.tablename IN (
      SELECT relname
      FROM pg_class c
      JOIN pg_partitioned_table p ON c.oid = p.partrelid
    )
  ORDER BY schemaname, tablename;
")

print(parents)

partitions <- dbGetQuery(U, "
  SELECT
    child.relname AS partition_name,
    parent.relname AS parent_table,
    ns_child.nspname AS schema_name
  FROM pg_inherits
  JOIN pg_class parent ON pg_inherits.inhparent = parent.oid
  JOIN pg_class child ON pg_inherits.inhrelid = child.oid
  JOIN pg_namespace ns_child ON child.relnamespace = ns_child.oid
  ORDER BY parent.relname, child.relname;
")

print(partitions)

dbListTables(U)

```

```{r test run eval=FALSE}
# get data
dsn = "PG:host=doyenne.csde.washington.edu dbname=nwi_smartmap user=aybloom password=csde4322 port=5432" 
f_tract_getdata_0one(year=2016, state='53', county = '077', verbose=TRUE)

## troubleshooting: 
dbExecute(U, 'DROP TABLE IF EXISTS tmp.tct')
dbExecute(U, "DROP TABLE IF EXISTS tmp.tract;")

# x_file <- st_read(glue('{tmp}/x.gpkg'))

## does not work!!
#cmd_test <- 'ogr2ogr -f PostgreSQL "PG:host=doyenne.csde.washington.edu dbname=nwi_smartmap user=aybloom password=csde4322 port=5432" #C:/Temp/4/RtmpUTYd0F/x.gpkg -nln tmp.tct nlt MULTIPOLYGON'
#system(cmd_test)

#cmd_test2 <- glue('ogr2ogr -f PostgreSQL "{dsn}" {tmp}\\x.gpkg x -lco geometry_name=geom_4326 -nln tct -lco schema=tmp -lco OVERWRITE=YES #-nlt multipolygon')
#system(cmd_test2)

## 10/22 -- updated the function to use st_write to write the table to tmp.tct in the dv

dbGetQuery(U, "SELECT table_schema, table_name FROM information_schema.tables WHERE table_schema='tmp';")

# check table
dbGetQuery(U, "SELECT * FROM tract.tract LIMIT 5;") # written above (final file)
dbGetQuery(U, "SELECT * FROM tmp.tract LIMIT 5;") ## written in the T() function
dbGetQuery(U, "SELECT * FROM tmp.tct LIMIT 5;") # written above (temp file)
```

```{r pull in all the data for 2016, WA state}
states_fips <- '53'
counties_fips <- fips_codes %>% dplyr::filter(state_code %in% states_fips)

# years 
years<-c(2016)

f_getdata_1all(startrun = 1, vrb = TRUE)
```

```{r view the tract.tract table data}
dbGetQuery(U, "SELECT * FROM tract.tract LIMIT 15;")

# Unique states present
states <- dbGetQuery(U, "SELECT DISTINCT statefp FROM tract.tract ORDER BY statefp;")
print(states)

# Total number of rows
nrows <- dbGetQuery(U, "SELECT COUNT(*) AS total FROM tract.tract;")
print(nrows$total) ## total should be 1458 (total census tracts in WA state)
```

## download water shapefiles
```{r set up tables and schema for water}
source('tools/area_water_download_postgis_partition.R')
# states and counties FIPS codes
states_fips <- '53'
counties_fips <- fips_codes %>% dplyr::filter(state_code %in% states_fips)

#update years: 
years<-c(2016)

dsn = "PG:host=doyenne.csde.washington.edu dbname=nwi_smartmap user=aybloom password=csde4322 port=5432" 

# create schemas
f_water_create_schema()

#create table for single tract
W(overwrite = TRUE)

# create partitions
f_water_create_partitions()

```

```{r test water download}
f_water_getdata_0one(year = 2016, state = '53', county = '033', verbose = TRUE)

# check table
dbGetQuery(U, "SELECT * FROM tmp.area_water LIMIT 10;")
dbGetQuery(U, "SELECT * FROM area_water.area_water LIMIT 10;")

area_water_tbl<-dbGetQuery(U, "SELECT * FROM area_water_state.area_water_2016_53 LIMIT 10;")
head(area_water_tbl)

```

## download all the water and put in partitioned tables
```{r download water for WA state for 2016}
states_fips <- '53'
counties_fips <- fips_codes %>% dplyr::filter(state_code %in% states_fips)

# years 
years<-c(2016)

f_water_getdata_1all(startrun = 1, vrb = TRUE)

# check table
area_water_tbl<-dbGetQuery(U, "SELECT * FROM area_water_county.area_water_2016_53_009")
head(area_water_tbl)

colnames(area_water_tbl
         )
```

## remove water from tract area shapefiles
```{r create nonwater partitions and schemas}
source('tools/tracts_erase_water.R')
## set list of years & states 
year_list <- c(2006, 2016)
state_abbr_list = c('wa', 'mn')

## create no_water schemas and partitions
f_nowater_create_schemas()
f_nowater_tract()
f_nowater_create_partitions()
```

## erase water
```{r test erase water download}
counties_df %<>% filter(state_code == '53')

# erase for King County table (county = 33)
f_erase_0one(i=17)



#con_str = "PG:host=db dbname=walk_db user=aybloom password=census port=5432"
con_str = "PG:host=doyenne.csde.washington.edu dbname=nwi_smartmap user=aybloom password=csde4322 port=5432" 


# check table - water removed
tract_with_water<-st_read(con_str, query = glue('SELECT * from tract.tract'))
tract_no_water_tbl<-st_read(con_str, query = glue('SELECT * from tract_nowater_state.tract_nowater_2016_53'))
state_nowater_tbl <- st_read(con_str, query = 'SELECT * from tract_nowater.tract_nowater')
water<-st_read(con_str, query = glue('SELECT * from area_water.area_water'))

tract_with_water[which(tract_with_water$tractce == '031800'),] %>% leaflet() %>% addTiles() %>% addPolygons()

# pick your tract
tract_sel <- tract_with_water %>% 
  filter(tractce == "031800")

# make sure everything is in same CRS
water <- st_transform(water, 4326)
tract_sel <- st_transform(tract_sel, 4326)

# spatial filter: keep only water polygons that intersect tract
water_in_tract <- st_intersection(water, tract_sel)

# get bounding box of the tract (or of water_in_tract)
bb <- st_bbox(tract_sel)

# plot water just in the tract:
leaflet() %>%
  addTiles() %>%
  addPolygons(data = water_in_tract, color = "blue", weight = 1, fillOpacity = 0.5) %>%
  addPolygons(data = tract_sel, color = "red", weight = 2, fill = FALSE)

tract_no_water_tbl[which(tract_no_water_tbl$tractce == '031800'),] %>% leaflet() %>% addTiles() %>% addPolygons()

#another tract with water:
tract_no_water_tbl[which(tract_no_water_tbl$tractce == '004100' & tract_no_water_tbl$countyfp == '033'),] %>% leaflet() %>% addTiles %>% addPolygons()

```

## erase all the water and put in partitioned no water tables
```{r erase water for WA state for 2016}
states_fips <- '53'
counties_fips <- fips_codes %>% dplyr::filter(state_code %in% states_fips)

# years 
years<-c(2016)

f_erase_1all(startnum = 1, verbose = TRUE)

## view a tract_nowater.tract_nowater table for county 033 for state 53
dbGetQuery(conn = U, 
           statement = 'SELECT * FROM tract_nowater_county.tract_nowater_2016_53_033 LIMIT 5')

```

## Download population counts
```{r download population counts}
source('tools/tracts_population_download.R')
## update states / years to WA 2016
# states and counties FIPS codes
states_fips <- "53"
counties_fips <- fips_codes %>% dplyr::filter(state_code %in% states_fips)

# years
#years <- c(2000, 2010:2020)
years<-2016

# these next create data frames enumerating the tables to be created
# year-state-county
counties_fips <- fips_codes %>%
  dplyr::filter(state_code %in% states_fips) %>%
  crossing(years)

## update counties_df
# tracts-years-states
counties_df <- counties_fips %>%
  select(years, state_code, county_code) %>%
  distinct() %>%
  mutate(tablename = str_c("tract_variables_county.tract_variables", years, state_code, county_code, sep = "_")) %>%
  # drop alaska
  filter(state_code != "02") %>%
  arrange(state_code, years, county_code) %>%
  mutate(rn = 1:nrow(.)) %>%
  select(rn, everything())

## create schemas/partitions:
f_pop_create_schemas()
f_pop_template()
f_pop_create_partitions()

```
```{r get King county data}
# get data for King County tracts: 

tract_table <- f_tractvar_0one(myYear = 2016, myState_fips = '53', myCounty_fips = '033', vrb = TRUE)

# view table
dbGetQuery(U, "SELECT * FROM tract_variables.tract_variables LIMIT 15;")

# check table
tract_tbl<-dbGetQuery(U, "SELECT * FROM tract_variables_county.tract_variables_2016_53_033 LIMIT 5")
head(tract_tbl)

colnames(tract_tbl)

```
```{r download all WA state 2016 population county data}
f_tractvar_1all(startnum = 1, verbose = TRUE)
```
```{r get population per cell}
source('tools/tract_population_per_cell.R')
counties_df %<>% filter(state_code == '53' & year == 2016)


# con_str = "PG:host=db dbname=walk_db user=aybloom password=census port=5432"
con_str = "PG:host=doyenne.csde.washington.edu dbname=nwi_smartmap user=aybloom password=csde4322 port=5432" 

f_merge_population_tract_nowater_1all(startnum = 1, 
                                      replace = TRUE, 
                                                  resolution_m = 100,
                                                  verbose = TRUE)

tract_nowater <- st_read(con_str, query = glue("SELECT * from tract_nowater.tract_nowater"))
head(tract_nowater)
```

## Rasterize the tract no water shapefiles and population counts and save to container
```{r rasterize tract files}
source('tools/tracts_rasterize_partition_data.R')
counties_df %<>% filter(state_code == '53' & year == 2016)
# test one
#f_rasterize_0one(conn = U, i = 1, resolution_m = 100, verbose = TRUE)

dsn = "PG:host=doyenne.csde.washington.edu dbname=nwi_smartmap user=aybloom password=csde4322 port=5432" 
# rasterize all data
f_tract_var_rasterize_1all(conn = U, resolution_m = 100, replace = FALSE)

## check a raster file
list.files('~')
tract_raster_land <- rast(glue('data/tract_raster_county/2016/53/tract_raster_land_area_2016_53_001.tif'))
tract_raster_pop <- rast('data/tract_raster_county/2016/53/tract_raster_population_2016_53_001.tif')
res(tract_raster_land)

plot(tract_raster_pop)
plot(tract_raster_land)
```
## combine tract-level raster files into a state file
```{r combine tract rasters}
source('tools/focal_area_functions.R')

# combine to state-level raster
f_combine_tract_land_raster(year = 2016, state = 53, 
                            county_filepath = 'data/tract_raster_county',
                            state_filepath = 'data/state_raster/land_area')
f_combine_tract_pop_raster(year = 2016, state = 53, 
                           county_filepath = 'data/tract_raster_county',
                           state_filepath = 'data/state_raster/population_density')

# test map the state raster
state_raster <- rast(glue('data/state_raster/land_area/2016/raw/state_raster_raw_2016_53.tif'))
plot(state_raster)

pop_raster <- rast(glue('data/state_raster/population_density/2016/raw/state_raster_raw_2016_53.tif'))
plot(pop_raster)
# check the resolution of the state raster
res(state_raster)
print(crs(state_raster))
res(pop_raster)
```

## Focal Sum for 1km circle area per 100m cell -- find the counts of cells (with land area vs. water) within 1 km. 
```{r focal sum stats}
source('tools/focal_area_functions.R')

# process land area focal sums -- find counts of land area for 1km, then convert to hectares (divide by 100) -
f_land_focus_processing(year = 2016, state_code = '53', 
                        radius_cir_crs_units = 1000 # radius of circle in the crs units (meters)
)

f_pop_focus_processing(year = 2016, state_code = '53', 
                       radius_cir_crs_units = 1000)


## state_raster_land_area_sum{year}_{state}.tif --> land area in meters per 1km circle
## state_raster_land_area_ha_{year}_{state_code}.tif --> land area in hectares per 1km circle 

## check the ha files
land_ha <- rast('data/state_raster/land_area/2016/ha/state_raster_land_area_ha_2016_53.tif')
pop_ha <- rast('data/state_raster/population_density/2016/ha/state_raster_land_area_ha_2016_53.tif')

plot(land_ha)
plot(pop_ha)

```

## Intersections & Edges ##
```{r download edges}
## update value for shutdown to pull from archives
GOVT_SHUTDOWN <- TRUE
source('tools/download_roads_edges.R')
## udpate where to download to: 
# where to download
dldir <- file.path("data/gisdata/edges")

download_edges(wa_fips = wa_fips, year = 2016)
```

```{r push edges to db}
#source('tools/push_edges_to_db.R')
source('tools/push_edges_to_pg.R')

dsn = "PG:host=doyenne.csde.washington.edu dbname=nwi_smartmap user=aybloom password=csde4322 port=5432"

years<-c(2016)
states_fips <- '53'
counties_fips <- counties_fips %>% filter(state_code == '53')

f_create_schema()
E()
f_partitions_year()
f_partitions_year_state()
f_partitions_year_state_county()

## check file / schema structures
dbGetQuery(U, "SELECT table_schema, table_name FROM information_schema.tables WHERE table_schema = 'edges' order by table_schema, table_name;")
 dbGetQuery(conn = U, statement = "SELECT table_schema, table_name FROM information_schema.tables where table_schema = 'edges_detail' order by table_schema, table_name;")

f_upsert_shapefiles(year = 2016, conn = U, states_fips = states_fips)

con_str = "PG:host=doyenne.csde.washington.edu dbname=nwi_smartmap user=aybloom password=csde4322 port=5432"
## error -- constraint violation
# look at tmp file: 
edges_tmp <- st_read(con_str, query = 'SELECT * FROM tmp.edge')
head(edges_tmp)

# look at edge file format
edges_file <- st_read(con_str, query = 'SELECT * FROM edges_state.edge_2016_53')
head(edges_file)

# chec, partition constraints
dbGetQuery(U, statement = "SELECT * FROM pg_constraint;")

dbGetQuery(U, statement = "SELECT * FROM pg_constraint c JOIN pg_class t ON c.conrelid = t.oid;")

dbGetQuery(U, "SELECT column_name, data_type FROM information_schema.columns WHERE table_name = 'edge_2016_53_001';")
dbGetQuery(U, "SELECT column_name, data_type FROM information_schema.columns WHERE table_schema = 'tmp';")


## snapper tool (snap each edge to the endpoints?)
f_snapper(conn = U)
f_index(geomfield = 'geom', conn = U)

edges_tbl <- st_read(con_str, query = 'SELECT * from edges_detail.edge_2016_wa_033')
head(edges_tbl)

## check for indices and gists: 
sql <- "SELECT schemaname, tablename, indexname, indexdef FROM pg_indexes WHERE schemaname IN ('edges', 'edges_detail', 'edges_state') ORDER BY schemaname, tablename, indexname;"
dbGetQuery(conn = U, statement = sql)

```

```{r create intersections}
source('tools/create_intersection_buffers.R')
build_endpoints(conn = U, year = 2016, state_abbr = 'wa')
## check endpoints table
build_intersections(conn = U, state_abbr = 'wa', year = 2016)
### view WA intersection/endpoints table: 
endpoints_tbl <- st_read(con_str, query = 'SELECT * from edges.endpoints_53_2016')
head(endpoints_tbl)
table(endpoints_tbl$mtfcc)

## plot the endpoints to verify: 
plot(endpoints_tbl)

build_intersections(conn = U, year = 2016, state_abbr = 'wa')

## check intersections table
intersections_tbl <- st_read(con_str, query = 'SELECT * from edges.intersections_3plus_53_2016')
intersections_tbl <- st_read(con_str, query = 'SELECT * from edges.intersections_3plus_53_2007')


head(intersections_tbl)

## tranform and map with leaflet
intersections <- st_transform(intersections_tbl, 4326)

##only plot for King county: 
tract_county_tbl <- st_read(con_str, query = 'SELECT * from tract_nowater_county.tract_nowater_2016_53_033')
king_county_shp <- st_union(tract_county_tbl)
king_county_shp <- st_as_sf(data.frame(geometry = king_county_shp))
county <- st_transform(king_county_shp, st_crs(intersections))
intersections_in_county <- st_intersection(intersections, county)

#map intersections
leaflet(intersections_in_county) %>%
  addTiles() %>%  # default OpenStreetMap tiles
  addCircleMarkers(
    radius = 3,                       # marker size
    color = "red",                     # marker color
    stroke = FALSE,                    # no border
    fillOpacity = 0.4,                 # transparency
    popup = ~paste("OID:", oid, "<br>",
                   "Points:", n_points, "<br>",
                   "Year:", year)
  )

## Look at some intersections: 
# intersections %>% filter(oid == '318076')
# edges_tbl %>% filter(tlid %in% c(187104515, 187110446, 187106283)) %>%
#   leaflet() %>% addTiles() %>%
#   addPolylines()
```

```{r rasterize intersections}
source('tools/rasterize_state_intersections.R')
con_str = "PG:host=doyenne.csde.washington.edu dbname=nwi_smartmap user=aybloom password=csde4322 port=5432"
f_rasterize_intersections_0one(con_str = con_str, 
                               conn = U, 
                               state_fips = '53', year = 2016, 
                               resolution_m = 100, verbose = TRUE, 
                               state_filepath = 'data/state_raster/intersections')

## preview raster data: 
wa_intersections_r <- rast('data/state_raster/intersections/2016/raw/state_raster_intersection_buffer_2016_53.tif')
head(wa_intersections_r)
plot(wa_intersections_r)

## get intersection density
f_get_intersection_density(con_str = con_str, 
                           conn = U, state_fips = '53', year = 2016)

## show intersection density raster
#preview intersection density - check max value
intersect_density <- rast('data/state_raster/intersections/2016/ha/state_raster_intersection_density_2016_53.tif')
plot(intersect_density)

```


## Business Density ####
```{r separate business raster}
source('tools/separate_wa_business_rasters.R')

#check raster
## preview raster data: 
wa_business_r <- rast('data/state_raster/business/2016/sum/state_raster_business_density_2016_53.tif')
head(wa_business_r)
plot(wa_business_r)


```
```{r get business density}
source('tools/get_business_density.R')
f_get_business_density(state_fip = '53', year = 2016, verbose = TRUE)
```

## create validation points & buffers
```{r create point buffer}
source('tools/validate_smartmap_measures.R')
f_create_validation_schema()
con_str = "PG:host=doyenne.csde.washington.edu dbname=nwi_smartmap user=aybloom password=csde4322 port=5432"
f_create_validation_points_df(state_fips = '53', con_str = con_str) ## update on the file to set different validation points
f_create_validation_buffers(conn = U, state_fips = '53')
f_validation_land_area(conn = U, year = 2016, state_fips = '53')

# view table
dbGetQuery(U, "SELECT * FROM validation.validation_points;")

con_str = "PG:host=doyenne.csde.washington.edu dbname=nwi_smartmap user=aybloom password=csde4322 port=5432"
val_points <- st_read(con_str, query = 'SELECT * from validation.validation_points')
val_buffer <- st_read(con_str, query = glue('SELECT * from validation.buffers'))
val_land <- st_read(con_str, query = ('SELECT * from validation.land_area'))
head(val_points)
head(val_buffer)
head(val_land)

val_points %>% leaflet() %>% addTiles() %>% addMarkers()

# set active geometry for buffer
st_geometry(val_buffer) <- val_buffer$geom_4326
st_geometry(val_land) <- val_land$geom_4326
# map points & buffers to check
val_buffer %>% 
  #dplyr::select(pid, geom_4326) %>% 
  leaflet() %>% addTiles() %>% addPolygons(data = val_buffer, color = 'blue', weight = 1, fillOpacity = 0.3) %>%
  addCircleMarkers(data = val_points, color = 'red', radius = 2, fillOpacity = 0.8, label = ~pid) %>%
  addPolygons(data = val_land, color = 'darkviolet', fillOpacity = 0.5)
```

## Pull data for validation points
```{r pull validation data}

validate_result <- f_validate_smartwalk(dsn = con_str, year = 2016, state_fips = '53')

# Show the table
DT::datatable(validate_result$data %>% st_drop_geometry() %>%
                dplyr::select(gid, smartmap_area_ha, buffer_area_ha, `% diff`, intersection_density_ha, pop_density_ha, bus_density_ha) %>% 
                mutate(across(where(is.numeric), ~round(., 2)))
              )

```

```{r show validate map}
# Show the map
validate_result$map
```
