---
title: "Example Walk Measure Map"
date: "2025-11-19"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
library(terra)
setwd('H:/projects/beh_nwi_smartmap')
source('tools/get_map_template_functions.R')
source('tools/validate_smartmap_measures.R')
con_str = "PG:host=doyenne.csde.washington.edu dbname=nwi_smartmap user=aybloom password=csde4322 port=5432"
```
## Set county and years
```{r example settings, echo = TRUE,eval = TRUE}
year_template <- 2016
year_one <- 2007
year_two <- 2024

county_name = 'King County'
```
```{r setup template, eval = TRUE}
invisible(
  capture.output(
    suppressMessages(
    suppressWarnings({
        shp_county_template <- get_county_template(county = 'King County', year = year_template)
    })
    )
    )
  )


```
```{r get cropped raster, eval = TRUE}
ha_cropped_rast_2007 <- crop_ha_files(year = year_one, 
                                 county_template = shp_county_template)

ha_cropped_rast_2024 <- crop_ha_files(year = 2024, 
                                      county_template = shp_county_template)

invisible(
  capture.output(
    suppressMessages(
    suppressWarnings({
king_template <- tigris::counties(state = 'WA') 
king_template <- king_template %>% 
  dplyr::filter(COUNTYFP == '033')
King_template_26910 <- st_transform(king_template, 26910)
    })
)
)
)

```
## Population density maps
```{r pop density maps}
pop_2007 <- ha_cropped_rast_2007[[1]]
pop_2024 <- ha_cropped_rast_2024[[1]]

pal <- colorNumeric(c('white', 'cyan4', "#0C2C84"), values(pop_2024),
  na.color = "transparent")

pop_2007_map<-leaflet() %>%
  addTiles() %>%
  addRasterImage(pop_2007, colors = pal, opacity = 0.75) %>%
  addLegend(pal = pal, values = values(pop_2007), title = 'Pop density , 2007')

pal <- colorNumeric(c('white', 'cyan4', "#0C2C84"), values(pop_2024),
  na.color = "transparent")

pop_2024_map<-leaflet() %>%
  addTiles() %>%
  addRasterImage(pop_2024, colors = pal, opacity = 0.75) %>%
  addLegend(pal = pal, values = values(pop_2007), title = 'Pop density , 2024')

```
### People per hectare (for 1km radius area)
```{r print pop maps, fig.width=800}
library(leafsync)
leafsync::sync(pop_2007_map, 
               pop_2024_map)
```
### Change in people per hactare (for 1km radius area)
```{r get change in pop density}
# get overlapping extents only

pop_2024_aligned <- terra::resample(pop_2024, pop_2007, method = 'bilinear')
pop_density_change = pop_2024_aligned - pop_2007
#pop_density_change$focal_sum[is.na(pop_density_change$focal_sum)] <- 0

shp_county_26910 <- st_transform(shp_county_template, st_crs(pop_density_change))
pop_density_change_mask <- mask(pop_density_change, shp_county_26910)

```
```{r map population change}
pop_density_change_mask <- as.numeric(pop_density_change_mask)
NAflag(pop_density_change_mask) <- NA
eps <- 0.01

pop_density_change_mask <- ifel(
  abs(pop_density_change_mask) <= eps,
  NA,
  pop_density_change_mask
)

max_pop <- global(pop_density_change_mask, fun = 'max', na.rm=TRUE)$max
pal <- colorNumeric(
  palette = "RdBu",
  domain  = c(-max_pop, max_pop), ## create symmetry
  na.color = "transparent"
)


leaflet() %>%
  addTiles() %>%
  addRasterImage(pop_density_change_mask, colors = pal, opacity = 0.75) %>%
  addLegend(pal = pal, 
            position = c('bottomright'), 
            values = values(pop_density_change_mask), title = 'Pop density change <br> (people per hectare), <br> 2007 to 2024')

```
### Change in people per hectare summary stats
```{r pop change summary stats, fig.width = 300}
library(terra)
library(dplyr)
library(knitr)
library(terra)

pop_density_change_mask <- mask(pop_density_change, shp_county_26910)
r <- pop_density_change_mask

stats <- c(
  min = round(global(r, min, na.rm = TRUE)[1,1],3),
  p10 = round(global(r, quantile, probs = 0.10, na.rm = TRUE)[1,1], 3),
  q1  = round(global(r, quantile, probs = 0.25, na.rm = TRUE)[1,1], 3),
  median = round(global(r, median, na.rm = TRUE)[1,1],3),
  mean = round(global(r, mean, na.rm = TRUE)[1,1],3),
  q3  = round(global(r, quantile, probs = 0.75, na.rm = TRUE)[1,1],3),
  p10 = round(global(r, quantile, probs = 0.90, na.rm = TRUE)[1,1], 3),
  max = round(global(r, max, na.rm = TRUE)[1,1],3)
)

stats.table <- data.frame(Value =  stats) %>% t() %>% as.data.frame()
colnames(stats.table) <- names(stats)
row.names(stats.table) <- NULL


knitr::kable(
  stats.table,
  caption = "Summary statistics, 2007 to 2024", digits = 3,
  align = 'c') %>%
  kable_classic() %>%
  kable_styling("striped", position = 'left', full_width = FALSE)
```

## Intersection density maps: 
```{r get intersection endpoints and lines for 2007, include = FALSE}
# get endpoints, filter to King county
endpoints_tbl_2007 <- st_read(con_str, query = 'SELECT * from edges.endpoints_53_2007')
endpoints_filter_2007 <- st_filter(endpoints_tbl_2007, King_template_26910)

# get edges and filter to local roads/paths 
edges_2007 <- st_read('H:/projects/beh_nwi_smartmap/gisdata/us_census/edges/fe_2007_53033_edges.shp')
edges_2007 <- edges_2007 %>%
  dplyr::filter(startsWith(MTFCC, "S")) %>% 
  dplyr::filter(!(MTFCC %in% c('S1100', 'S1500', 'S1630', 'S1640', 'S1730', 'S1750', 'S1780')))

# get intersections (3+ end/start points of lines)
intersections_3plus_tbl_2007 <- st_read(con_str, query = 'SELECT * from edges.intersections_3plus_53_2007') %>% 
  dplyr::select(-geom_snap_buffer_26910)
intersections_filter_2007 <- st_filter(intersections_3plus_tbl_2007, King_template_26910) 

# transform for leaflet map
endpoints_4326_2007 <- st_transform(endpoints_filter_2007, 4326)
edges_4326_2007 <- st_transform(edges_2007, 4326)
intersections_4326_2007 <- st_transform(intersections_filter_2007, 4326)

```
```{r get intersection endpoints and lines for 2024, include = FALSE}
# get endpoints, filter to King county
endpoints_tbl_2024 <- st_read(con_str, query = 'SELECT * from edges.endpoints_53_2024')
endpoints_filter_2024 <- st_filter(endpoints_tbl_2024, King_template_26910)

# get edges and filter to local roads/paths 
edges_2024 <- st_read('H:/projects/beh_nwi_smartmap/gisdata/us_census/edges/tl_2024_53033_edges.shp')
edges_2024 <- edges_2024 %>%
  dplyr::filter(startsWith(MTFCC, "S")) %>% 
  dplyr::filter(!(MTFCC %in% c('S1100', 'S1500', 'S1630', 'S1640', 'S1730', 'S1750', 'S1780')))

# get intersections (3+ end/start points of lines)
intersections_3plus_tbl_2024 <- st_read(con_str, query = 'SELECT * from edges.intersections_3plus_53_2024') %>% 
  dplyr::select(-geom_snap_buffer_26910)
intersections_filter_2024 <- st_filter(intersections_3plus_tbl_2024, King_template_26910) 

# transform for leaflet map
endpoints_4326_2024 <- st_transform(endpoints_filter_2024, 4326)
edges_4326_2024 <- st_transform(edges_2024, 4326)
intersections_4326_2024 <- st_transform(intersections_filter_2024, 4326)

```
```{r intersection density maps, eval = TRUE}
library(leaflet.extras2)
intersection_2007 <- ha_cropped_rast_2007[[2]]
intersection_2024 <- ha_cropped_rast_2024[[2]]

intersection_2007_4326<-project(intersection_2007, 'EPSG:4326')
intersection_2024_4326<-project(intersection_2024, 'EPSG:4326')


pal <- colorNumeric(c('white', 'cyan4', "#0C2C84"), 
                    values(intersection_2007_4326),
                    na.color = "transparent")

intersection_density_2007 <- leaflet() %>%
  addTiles() %>%
# ## add roads & paths
        addPolylines(
          data = edges_4326_2007,
          color = 'steelblue',
          weight = 3,
          opacity= 0.5,
          popup = ~paste('MTFCC: ', MTFCC),
          group = 'intersections'
  ) %>%
  addCircleMarkers(
    data = intersections_4326_2007,
          radius = 10,
          fillColor = 'red',
          fillOpacity = 0.7,
          stroke = FALSE,
          popup = ~paste0("tlids: ", tlid),
          clusterOptions = markerClusterOptions(),
          group = 'intersections'
) %>%
  addRasterImage(intersection_2007_4326, colors = pal, opacity = 0.75) %>% 
  groupOptions(group = 'intersections', zoom = 14) %>%
  addLegend(pal = pal, values = values(intersection_2007_4326), title = 'Intersections<br>
            (per hectare),<br>2007')


intersection_density_2024 <- leaflet() %>%
  #addTiles(group = "Modern (Live)") %>%
  addTiles()%>%
  addRasterImage(intersection_2024_4326, colors = pal, opacity = 0.75) %>%
  
   addPolylines(
          data = edges_4326_2024, 
          color = 'steelblue', 
          weight = 3, 
          opacity= 0.5,
          popup = ~paste('MTFCC: ', MTFCC),
          group = 'intersections'
                     ) %>%
  addCircleMarkers(
    data = intersections_4326_2024,
          radius = 10,
          fillColor = 'red',
          fillOpacity = 0.7,
          stroke = FALSE,
          popup = ~paste0("tlids: ", tlid),
          clusterOptions = markerClusterOptions(),
          group = 'intersections') %>%
    groupOptions(group = 'intersections', zoom = 14) %>%
  addLegend(pal = pal, values = values(intersection_2024_4326), title = 'Intersection <br> 
            (per hectare), <br> 
            2024')

```
### Intersections per hectare (for 1km radial area around cell)*
```{r print intersection maps, eval=TRUE, fig.width=8 }
library(leafsync)
leafsync::sync(
                intersection_density_2007, 
               intersection_density_2024, 
               ncol = 1)
```

### Change in intersections per hectare (for 1km radial area), 2007 to 2024 
```{r get change in intersection density, eval = TRUE}
# get overlapping extents only
intersection_2024_aligned <- terra::resample(intersection_2024_4326, intersection_2007_4326, method = 'bilinear')
intersection_density_change = intersection_2024_aligned - intersection_2007_4326

```
```{r map intersection change, eval = TRUE}
intersection_density_change_mask <- as.numeric(intersection_density_change)
NAflag(intersection_density_change_mask) <- NA
eps <- 0.001

# intersection_density_change_mask <- ifelse(
#   abs(intersection_density_change_mask) <= eps,
#   NA,
#   intersection_density_change_mask
# )

max_intersect <- global(abs(intersection_density_change_mask), fun = 'max', na.rm=TRUE)$max

pal <- colorNumeric(
  palette = "RdBu",
  domain  = c(-max_intersect, max_intersect), ## create symmetry
  na.color = "transparent"
)

#writeRaster(intersection_density_change_mask, 'H:/projects/beh_nwi_smartmap/intersection_qa_files/change_intersection_2007_2024_C.tiff', overwrite=TRUE)

intersection_change_map <- leaflet() %>%
  addTiles() %>%
  addRasterImage(intersection_density_change_mask, colors = pal, opacity = 0.75) %>%
  addLegend(pal = pal, 
            position = c('bottomright'), 
            values = values(intersection_density_change_mask), title = 'Change in intersection density, <br>2007 to 2024')
```

```{r print intersection change map}
intersection_change_map
```

### Change in intersection density summary stats
```{r intersection change summary stats}
library(terra)
library(dplyr)
library(knitr)
library(terra)

r <- intersection_density_change

stats <- c(
  min = round(global(r, min, na.rm = TRUE)[1,1],3),
  p10 = round(global(r, quantile, probs = 0.10, na.rm=TRUE)[1,1], 3),
  q1  = round(global(r, quantile, probs = 0.25, na.rm = TRUE)[1,1], 3),
  median = round(global(r, median, na.rm = TRUE)[1,1],3),
  mean = round(global(r, mean, na.rm = TRUE)[1,1],3),
  q3  = round(global(r, quantile, probs = 0.75, na.rm = TRUE)[1,1],3),
  p95 = round(global(r, quantile, probs = 0.90, na.rm = TRUE)[1,1], 3),
  max = round(global(r, max, na.rm = TRUE)[1,1],3)
)


stats.table <- data.frame(Value =  stats) %>% t() %>% as.data.frame()
colnames(stats.table) <- names(stats)
row.names(stats.table) <- NULL

knitr::kable(
  stats.table, 
  caption = "Summary statistics, 2007 to 2024", digits = 3) %>%
  kable_classic() %>%
kableExtra::kable_styling(full_width = F)

```

## Business density maps
```{r  business density maps}
 business_2007 <- ha_cropped_rast_2007[[3]]
 business_2024 <- ha_cropped_rast_2024[[3]]

pal <- colorNumeric(c('white', 'cyan4', "#0C2C84"), values(business_2007),
  na.color = "transparent")

business_2007_map <- leaflet() %>%
  addTiles() %>%
  addRasterImage( business_2007, colors = pal, opacity = 0.75) %>%
  addLegend(pal = pal, values = values( business_2007), title = ' business density , 2007')

 pal <- colorNumeric(c('white', 'cyan4', "#0C2C84"), values(business_2024),
   na.color = "transparent")

business_2024_map<-leaflet() %>%
  addTiles() %>%
  addRasterImage( business_2024, colors = pal, opacity = 0.75) %>%
  addLegend(pal = pal, values = values( business_2007), title = ' business density , 2024')

```

### Businessess per hectare (for 1km radial area around cell)
```{r print business maps}
library(leafsync)
leafsync::sync(business_2007_map, 
               business_2024_map)
```

### Change in business density (businessess per hectare for 1km radial area), 2007 to 2024
```{r get change in  business density}
# get overlapping extents only

 business_2024_aligned <- terra::resample( business_2024,  business_2007, method = 'bilinear')
 business_density_change =  business_2024_aligned -  business_2007
# business_density_change$focal_sum[is.na( business_density_change$focal_sum)] <- 0

shp_county_26910 <- st_transform(shp_county_template, st_crs( business_density_change))
 business_density_change_mask <- mask( business_density_change, shp_county_26910)

```
```{r map business change}
#business_density_change_mask[business_density_change_mask == 0] <- NA

# Get top quartile for maping values
q90 <- global(business_density_change_mask, fun = quantile, probs = 0.90, na.rm=TRUE)[1, 1]
business_change_top <- business_density_change_mask
business_change_top[business_change_top < q90] <- NA # remove values < top 10%
business_change_top_pts <- as.points(business_change_top, na.rm=TRUE)
names(business_change_top_pts)[1] <- 'value'
business_change_pts_top_4326 <- project(business_change_top_pts, 'EPSG:4326')

business_density_change_mask_4326 <- project(business_density_change_mask, 'EPSG:4326')
#writeRaster(business_density_change_mask, 
#'H:/projects/beh_nwi_smartmap/intersection_qa_files/change_business_density_2007_2024.tiff', overwrite=TRUE)

pal <- colorNumeric(c('darkred', 'white', 'cyan4'), values( business_density_change_mask_4326),
  na.color = "transparent")

leaflet() %>%
  addTiles() %>%
  #  addCircleMarkers(
  #    data = business_change_top_pts,
  #   radius = 3,
  #   fillOpacity = 0.7,
  #   stroke = FALSE,
  #   popup = ~paste("Value:", round(value, 2))
  # ) %>%
  addRasterImage( business_density_change_mask_4326, colors = pal, opacity = 0.75) %>%
  addLegend(pal = pal, 
            position = c('bottomright'), 
            values = values( business_density_change_mask), title = 'Change in  business density, <br> 2007 to 2024')
```

### Change in business density summary stats
```{r business change summary stats}
library(terra)
library(dplyr)
library(knitr)

library(terra)

r <- business_density_change_mask

stats <- c(
  min = round(global(r, min, na.rm = TRUE)[1,1],3),
  p10 = round(global(r, quantile, probs = 0.10, na.rm=TRUE)[1,1], 3),
  q1  = round(global(r, quantile, probs = 0.25, na.rm = TRUE)[1,1], 3),
  median = round(global(r, median, na.rm = TRUE)[1,1],3),
  mean = round(global(r, mean, na.rm = TRUE)[1,1],3),
  q3  = round(global(r, quantile, probs = 0.75, na.rm = TRUE)[1,1],3),
  p95 = round(global(r, quantile, probs = 0.90, na.rm = TRUE)[1,1], 3),
  max = round(global(r, max, na.rm = TRUE)[1,1],3)
)


stats.table <- data.frame(Value =  stats) %>% t() %>% as.data.frame()
colnames(stats.table) <- names(stats)
row.names(stats.table) <- NULL

knitr::kable(
  stats.table, 
  caption = "Summary statistics, 2007 to 2024", digits = 3) %>%
  kable_classic() %>% 
  kable_styling(full_width = F)

```

## Composite Score maps: 